String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-25 2023-1-19/sections/introduction.tex
intro

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs for at least twice and at most four times and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that Kleene star and Kleene plus operator can be seen as special cases of counting operator, for instance, $a^*$ and $a^+$ are equivalent to $a^{\{0,\infty\}}$ and $a^{\{1,\infty\}}$.
Counting is also a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in over 70\% of the Python projects, while the other forms of counting occur in over 20\% of them.

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 60 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 40\}} (\Sigma \setminus b)^{\{5, 40\}} (\Sigma \setminus c)^{\{5, 40\}} \wedge x \in \Sigma^* c \wedge |x| > 80.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with $c$, and the length of $x$ is greater than $80$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 80$ and the counting upper bound $40$ in  both $(\Sigma \setminus a)^{\{5, 40\}}$ and $(\Sigma \setminus b)^{\{5, 40\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c$ requires that $x$ has to end with $c$.

@@

\paragraph*{Contribution} xxx

\paragraph*{Related work} xx

\paragraph*{Organization} xxx

@@

@@

The rest of the paper is structured as follows: Section \ref{sec:pre} introduces the preliminaries. Section \ref{sec:overview} outlines the overall algorithm. Section \ref{sec:automaton} illustrates how to construct an automaton from a regular expression. Section \ref{sec:algorithm} shows how to solve a string constraint with repetition. Section \ref{sec:implementation} presents the experimental results. Section \ref{sec:conclu} concludes the paper and discusses future work.

>>>>

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs for at least twice and at most four times and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that Kleene star and Kleene plus operator can be seen as special cases of counting operator, for instance, $a^*$ and $a^+$ are equivalent to $a^{\{0,\infty\}}$ and $a^{\{1,\infty\}}$.
Counting is also a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in more than 70\% of 1,204 Python projects, while the other forms of counting occur in more than 20\% of them.  Therefore, an efficient analysis of string manipulating programs requires efficient solvers for string constraints that contain at least regular expressions with counting and string-length function.

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{5, 60\}} \wedge x \in \Sigma^* c \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with $c$, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{5, 60\}}$ and $(\Sigma \setminus b)^{\{5, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c$ requires that $x$ has to end with $c$.

A typical way of string constraint solvers to deal with regex-counting is to unfold regular expressions with counting into those \emph{without} counting by utilizing the concatenation operator. For instance, $a^{\{0, 4\}}$ is unfolded into $\varepsilon + a + aa + aaa + aaaa$. Since the unfolding incurs an exponential blow-up on the constraint sizes (where the integer constants are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

@@

\paragraph*{Contribution} In this work, we focus on the class of string constraints with regular membership and string-length function, where the counting operators may occur in regular expressions (called RECL for brevity). We propose an automata-theoretical approach for solving RECL constraints. Our main idea is to represent the counting operators by registers in automata, instead of unfolding them explicitly.  Our approach is built upon an automata model called cost-enriched finite automata (CEFA) that was proposed in~\cite{atva2020}. Moreover, the string-length function can also be modeled by registers in CEFA. Therefore, the satisfiability of RECL constraints is reduced to the nonemptiness problem of CEFA.

According to the results from~ \cite{atva2020}, a linear integer arithmetic (LIA) formula can be computed to represent the potential values of registers in CEFA. Thus, the nonemptiness of CEFA can be reduced to the satisfiability of LIA formulas, which is then tackled by off-the-shelf SMT solvers for LIA. 
%
Because the sizes of LIA formulas depend much on the sizes of CEFA, it is beneficial to reduce the sizes of CEFA, in order to achieve a good performance. As a result, we propose various techniques to reduce the state space, the number of transitions, and the number of registers for CEFA. Furthermore, since the computation of LIA formulas from CEFA is relatively expensive, we also propose a method of searching for strings that are accepted by CEFA, based on under approximations, before constructing the LIA formulas.

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting into a new version of OSTRICH, called $\ostrichrecl$ for convenience.  
Furthermore, we utilize a collection of benchmark suites, which comprise 49,379 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contains sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 49,128 out of 49,379 instances, while the state-of-the-art best solver Z3str3RE solves only 47,407 instances, with a timeout of 60 seconds. Moreover, $\ostrichrecl$ and Z3str3RE spend almost the same amount of time in average, that is, 2.5 seconds, in solving each instance. In comparison, CVC5, Z3str3, and Z3seq solve only 44,001, 36,320, 46,813 instances respectively, with an average time 7.8, 7.7, 3.8 seconds per instance.

\paragraph*{Related work} \zhilin{to be cleaned} To handle the problems in software verification reasoning about strings and integers, string theories with regular membership predicate and linear length constraints are raised. String solvers such as Z3Str3\cite{z3str3}, Z3Seq\cite{z3seq}, BD+23\cite{BD+23}, CVC4\cite{cvc4}, CVC5\cite{cvc5}, (Z3-)Trau\cite{trau}\cite{z3trau}, Ostrich(+)\cite{ostrich}\cite{atva2020}, Sloth\cite{sloth}, S3\cite{s3}, and ABC\cite{abc} are developed on these theories. Although string theory is easily
undecidable\cite{undecidable_1}\cite{undecidable_2}, the satisfiability problems for string constraints of regular expressions, linear integer arithmetic, and the string length is decidable\cite{theory_BD+23}. There are mainly two strategies for solving string constraints: (1) DPLL(T)-based\cite{dpll_t} string solvers\cite{z3str3}\cite{cvc5} apply heuristic derivation rules to unwind the regular expression gradually and identify several classes of simplification techniques for efficiency. (2) Automata-based string solvers\cite{trau}\cite{z3trau}\cite{ostrich}\cite{atva2020}\cite{BD+23} construct a finite automaton to model the regular expression and use the emptiness problem of the automaton (under linear integer arithmetic) to decide the satisfiability of string constraints.

\paragraph*{Organization} \zhilin{to be cleaned} The rest of the paper is structured as follows: Section \ref{sec:pre} introduces the preliminaries. Section \ref{sec:overview} outlines the overall algorithm. Section \ref{sec:automaton} illustrates how to construct an automaton from a regular expression. Section \ref{sec:algorithm} shows how to solve a string constraint with repetition. Section \ref{sec:implementation} presents the experimental results. Section \ref{sec:conclu} concludes the paper and discusses future work.

@@

%%%%%%%%%%%%%%%%%%%%%%% original texts by Denghang %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%% original texts by Denghang %%%%%%%%%%%%%%%%%
\hide{

@@

The rest of the paper is structured as follows: Section \ref{sec:pre} introduces the preliminaries. Section \ref{sec:overview} outlines the overall algorithm. Section \ref{sec:automaton} illustrates how to construct an automaton from a regular expression. Section \ref{sec:algorithm} shows how to solve a string constraint with repetition. Section \ref{sec:implementation} presents the experimental results. Section \ref{sec:conclu} concludes the paper and discusses future work.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
